\bf The forestecology R package for fitting and assessing models of interspecies competition on the growth of trees

### 2021-02-22 etc/paper/forestecology.tex
Rewrite "permutation test" and "spatial cross-validation" sections

\if0\blind
{
  \title{\bf The \texttt{forestecology} R package for modeling interspecies
competition between trees}

  \author{
        Albert Y. Kim \thanks{Albert Y. Kim is Assistant Professor, Statistical \& Data Sciences,
Smith College, Northampton, MA 01063 (e-mail:
\href{mailto:akim04@smith.edu}{\nolinkurl{akim04@smith.edu}}).} \\
    Program in Statistical \& Data Sciences, Smith College\\
     and \\     David Allen \\
    Biology Department, Middlebury College\\
     and \\     Simon P. Couch \\
    Mathematics Department, Reed College\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The \texttt{forestecology} R package for modeling interspecies
competition between trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Move abstract below here after completed.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, competition, R, Rstats, tidyverse, sf, cross-validation,
\vfill

@@

\hypertarget{abstract-350-words}{%
\section*{Abstract (350 words)}\label{abstract-350-words}}
\addcontentsline{toc}{section}{Abstract (350 words)}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  When modeling growth of trees forest ecologists often incorporate the
  effect of interspecies competition. Many such models are based on a
  neighborhood effect assumption whereby all trees within a fixed
  distance of all focal trees are considered competitors. Methods are
  needed to evaluate the effect of interspecies competition and to
  assess their quality.
\item
  We present the \texttt{forestecology} package providing methods for
  both 1) evaluating the out-of-sample performance of our model using
  spatial-crossvaliation and 2) testing a null hypothesis of of no
  impact of competitor species' identity on the growth of trees using a
  permutation test. We implement a class and methods using R's S3
  object-oriented system, for a specific linear, Bayesian neighborhood
  competition model of tree growth.
\item
  We demonstrate the package's functions using data from the Smithsonian
  Conservation Biology Institute's large forest dynamics plot, part of
  the ForestGEO network of reseach sites. Given ForestGEO's data
  collection protocols and data formatting standards, the package
  cross-compatibility of code. We show both that 1) competitor species
  identity matters and 2) that not spatially cross-validating leads to
  error estimates that are overly optimistic.
\item
  The package follows \texttt{tidyverse}-like structure whereby
  verb-named functions can be modularly ``piped'' in sequence to
  intuitively display the sequence of steps of analysis from start to
  finish. Additionally, most inputs/outputs of functions assume an are
  of \texttt{sf} class from the simple features package, thereby
  facilitating all wrangling and visualization of geospatial data.
  Lastly, even though our package is currently limited to one specific
  model, the package is setup such that it can be easily extended to
  other models.
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute the growth of trees based on census data.
\item
  Add spatial information:

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    Define buffer region trees.
  \item
    Add spatial cross-validation block information.
  \end{enumerate}
\item
  Identify all focal and corresponding competitor trees.
\item
  Fit model and make predictions.
\end{enumerate}

@@

\hypertarget{compute-the-growth-of-trees-based-on-census-data}{%
\subsection{Compute the growth of trees based on census
data}\label{compute-the-growth-of-trees-based-on-census-data}}

@@

\caption{Compute growth of trees based on census data from the Smithsonian Conservation Biology Institute (SCBI).}\label{fig:scbi-trees}
\end{figure}

\hypertarget{spatial-information}{%
\subsection{Add spatial information}\label{spatial-information}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf_text}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{label =}\NormalTok{ foldID, }\DataTypeTok{col =}\NormalTok{ buffer)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ blocks_scbi, }\DataTypeTok{fill =} \StringTok{"transparent"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\caption{Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region will be the focal trees whose growth will modeled.}\label{fig:scbi-spatial-information}
\end{figure}

\hypertarget{focal-vs-comp}{%
\subsection{Identify all focal and corresponding competitor
trees}\label{focal-vs-comp}}

@@

Using the \texttt{unnest()} function from the \texttt{tidyr} package, we
can flatten list-column into regular columns. We observe that for the
same focal tree with DBH equal to 13.65cm, we have information on all 20
competitor trees whose \texttt{dist} distance to the focal tree is less
than or equal to 7.5, including their unique tree-stem ID number, their
species, and their basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) where \(DBH\) is the value
from the earlier census in cm. Saving our focal versus competitor
information in list-column minimizes redundancy since we do not repeat
information on the focal tree 20 times.

@@

\caption{Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}

\hypertarget{model-fit-predict}{%
\subsection{Fit model and make predictions}\label{model-fit-predict}}

@@

Lastly, the generic for \texttt{ggplot2::autoplot()} allows us to

@@

\caption{Fit model: Posterior distributions of all parameters for three species.}\label{fig:scbi-posterior-viz}
\end{figure}

@@

HEY BERT PICK IT UP HERE

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_bayes_lm_scbi <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{comp_bayes_lm}\NormalTok{(}\DataTypeTok{prior_param =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{run_shuffle =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{growth_hat =} \KeywordTok{predict}\NormalTok{(comp_bayes_lm_scbi, }\DataTypeTok{newdata =}\NormalTok{ focal_vs_comp_scbi)}
\NormalTok{  )}

\NormalTok{model_rmse <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse}
\end{Highlighting}
\end{Shaded}

@@

TODO: Do we talk about
\texttt{add\_buffer\_variable(direction\ =\ "out")} when each fold gets
its turn being the training data?

The model fits and predictions in Section \ref{model-fit-predict} all
suffer from a common failing: they use the same data to both fit the
model and to assess the model's performance using the RMSE. As argued by
\citet{roberts_cross-validation_2017}, this can lead to overly
optimistic assessments of model quality as the models can be overfit, in
particular in situations where spatial-autocorrelation is present. To
mitigate the effects of such overfitting, we use a spatially block
cross-validation algorithm implemented in the \texttt{run\_cv()}. This
function at its core uses the same model fitting implemented in the
\texttt{fit\_bayesian\_model()} function, however trains the model on
\(k-1\) spatial folds of the train and returns fitted values for the
test data. Recall that the spatial blocking scheme wass encoded in
Section \ref{spatial-information}.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\CommentTok{## [1] 0.1402209}
\end{Highlighting}
\end{Shaded}

@@

>>>>

\if0\blind
{
  \title{\bf The forestecology R package for fitting and assessing models of
interspecies competition on the growth of trees}

  \author{
        Albert Y. Kim \\
    Program in Statistical \& Data Sciences, Smith College\\
     and \\     David Allen \\
    Biology Department, Middlebury College\\
     and \\     Simon P. Couch \\
    Mathematics Department, Reed College\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The forestecology R package for fitting and assessing models of
interspecies competition on the growth of trees}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
1. When modeling growth of trees forest ecologists often incorporate the
effect of interspecies competition. Many such models are based on a
neighborhood effect assumption whereby all trees within a fixed distance
of all focal trees are considered competitors. Methods are needed to
evaluate the effect of interspecies competition and to assess their
quality.\\
2. We present the \texttt{forestecology} package providing methods for
both 1) evaluating the out-of-sample performance of our model using
spatial-crossvaliation and 2) testing a null hypothesis of of no impact
of competitor species' identity on the growth of trees using a
permutation test. We implement a class and methods using R's S3
object-oriented system, for a specific linear, Bayesian neighborhood
competition model of tree growth.\\
3. We demonstrate the package's functions using data from the
Smithsonian Conservation Biology Institute's large forest dynamics plot,
part of the ForestGEO network of reseach sites. Given ForestGEO's data
collection protocols and data formatting standards, the package
cross-compatibility of code. We show both that 1) competitor species
identity matters and 2) that not spatially cross-validating leads to
error estimates that are overly optimistic.\\
4. The package follows \texttt{tidyverse}-like structure whereby
verb-named functions can be modularly ``piped'' in sequence to
intuitively display the sequence of steps of analysis from start to
finish. Additionally, most inputs/outputs of functions assume an are of
\texttt{sf} class from the simple features package, thereby facilitating
all wrangling and visualization of geospatial data. Lastly, even though
our package is currently limited to one specific model, the package is
setup such that it can be easily extended to other models.
\end{abstract}

\noindent%
{\it Keywords:} forest ecology, competition, R, Rstats, tidyverse, sf, cross-validation,
spatial statistics
\vfill

@@

@@

\begin{itemize}
\tightlist
\item
  Step 1: Compute the growth of trees based on census data.
\item
  Step 2: Add spatial information:

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Define buffer region trees.
  \item
    Add spatial cross-validation block information.
  \end{enumerate}
\item
  Step 3: Identify all focal and corresponding competitor trees.
\item
  Step 4: Fit model, as well as:

  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    Compute fitted/predicted values.
  \item
    Visualize posterior distributions.
  \end{enumerate}
\end{itemize}

@@

\hypertarget{step-1-compute-the-growth-of-trees-based-on-census-data}{%
\subsection{Step 1: Compute the growth of trees based on census
data}\label{step-1-compute-the-growth-of-trees-based-on-census-data}}

@@

\caption{Step 1 - Compute growth of trees based on census data: All trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot.}\label{fig:scbi-trees}
\end{figure}

\hypertarget{spatial-information}{%
\subsection{Step 2: Add spatial information}\label{spatial-information}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ blocks_scbi, }\DataTypeTok{fill =} \StringTok{"transparent"}\NormalTok{, }\DataTypeTok{linetype =} \StringTok{"dashed"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf_text}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ growth_scbi }\OperatorTok{%>%}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\DecValTok{1000}\NormalTok{), }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{label =}\NormalTok{ foldID, }\DataTypeTok{col =}\NormalTok{ buffer))}
\end{Highlighting}
\end{Shaded}

@@

\caption{Step 2 - Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region will be the focal trees whose growth will modeled.}\label{fig:scbi-spatial-information}
\end{figure}

\hypertarget{focal-vs-comp}{%
\subsection{Step 3: Identify all focal and corresponding competitor
trees}\label{focal-vs-comp}}

@@

Using the \texttt{unnest()} function from the \texttt{tidyr} package, we
can flatten list-column into regular columns. We observe that for the
same focal tree with DBH equal to 13.65cm, we have information on all 20
competitor trees whose \texttt{dist} distance to the focal tree is less
than or equal to 7.5, including their unique tree-stem ID number, their
species, and their basal area (in m\(^2\)) calculated as
\(\frac{\pi \times (\text{DBH/2})^2}{10000}\) where \(DBH\) is the value
from the earlier census in cm. Saving our focal versus competitor
information in list-column minimizes redundancy since we do not repeat
information on the focal tree 20 times. The spatial distribution of
these trees is visualized in Figure \ref{fig:scbi-focal-vs-comp-map}:
the dashed circle extends 7.5 m away from the focal tree while all 20
competitor trees are within this circle.

@@

\caption{Step 3 - Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4.}\label{fig:scbi-focal-vs-comp-map}
\end{figure}

\hypertarget{model-fit-predict}{%
\subsection{Step 4: Fit model}\label{model-fit-predict}}

@@

Lastly, the generic for \texttt{ggplot2::autoplot()} allows us to plot
the posterior distribution of all parameters in Figure
\ref{fig:scbi-posterior-viz}.

@@

\caption{Step 4 - Fit model: Posterior distributions of all parameters for three species.}\label{fig:scbi-posterior-viz}
\end{figure}

@@

We use the four steps of our analysis sequence we answer the first of
our two questions: does competitor species identity matter when modeling
the growth of focal trees? We answer this via a permutation test: Under
a null hypothesis where competitor species identity does not matter we
can permute this variable within each focal tree, compute the RMSE (the
test statistic of interest), repeat this process several times to
construct a null distribution of RMSE, and compare the observed RMSE to
assess significance. Going back to our example in Section
\ref{focal-vs-comp} inspecting all 20 competitor trees to focal tree
with \texttt{focal\_ID} 4, the permutation test involves randomly
resampling the \texttt{comp\_sp} variable with replacement, leaving all
other competitor variables intact. The resampling with replacement is
nested within each focal tree in order to preserve the neighborhood
structure of our competitor model. To run the permutation test, we use
the same code as in Section \ref{model-fit-predict}, but adding a
\texttt{run\_shuffle\ =\ TRUE} argument to \texttt{comp\_bayes\_lm()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{comp_bayes_lm_scbi_shuffle <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{comp_bayes_lm}\NormalTok{(}\DataTypeTok{prior_param =} \OtherTok{NULL}\NormalTok{, }\DataTypeTok{run_shuffle =} \OtherTok{TRUE}\NormalTok{)}

\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{growth_hat_shuffle =} \KeywordTok{predict}\NormalTok{(comp_bayes_lm_scbi_shuffle, }\DataTypeTok{newdata =}\NormalTok{ focal_vs_comp_scbi))}

\NormalTok{model_rmse_shuffle <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat_shuffle) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse_shuffle}
\CommentTok{## [1] 0.131693}
\end{Highlighting}
\end{Shaded}

The resulting RMSE of 0.131693 based on the permutation test is larger
than the earlier RMSE of 0.1281398, suggesting that models that do
incorporate competitor species identity better fit the data. We conduct
a fuller simulation in Section below.

@@

Using the four steps of our analysis sequence again, we answer the
second of our two questions: how can we obtain an accurate estimate of
model performance/error? The model fits and predictions in Section
\ref{model-fit-predict} all suffer from a common failing: they use the
same data to both fit the model and to assess the model's performance
using the RMSE. As argued by \citet{roberts_cross-validation_2017}, this
can lead to overly optimistic assessments of model quality as the models
can be overfit, in particular in situations where
spatial-autocorrelation is present. To mitigate the effects of such
overfitting, we use a spatially block cross-validation algorithm.

To this end, we use the \texttt{foldID} variable defined in Section
\ref{spatial-information} whereby all focal trees are assigned to one of
4 spatially contiguous blocks that act as folds in our cross-validation
routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic}
presents a schematic illustrating this scheme for fold 1 as the test set
and folds 2, 3, and 4 as the training sets. We fit the model to all
focal trees in the training set, apply the model to all focal trees in
the test set to compute fitted/predicted values, and compute the RMSE of
the observed versus predicted growths. We repeat this procedure 3 more
times with each of the three remaining folds acting as the test set.
Furthermore, in order to maintain spatial independence between the test
and training set, a buffer that extend outwards from the boundary of the
test set is computed; all trees falling within this buffer will not be
part of the training set.

{\centering \includegraphics[width=0.66\linewidth]{Figures/scbi-spatial-cross-validation-schematic-1}

\caption{Schematic of spatial cross-validation: Using the k = 1 fold as the test set, assigning each focal tree to training set, test set, and buffer.}\label{fig:scbi-spatial-cross-validation-schematic}
\end{figure}

This algorithm is implemented in the \texttt{run\_cv()} function, which
acts as a wrapper function to the \texttt{comp\_bayes\_lm()} function
that fits the model and the \texttt{predict()} generic that returns
fitted values.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{model_rmse_cv <-}\StringTok{ }\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{rmse}\NormalTok{(}\DataTypeTok{truth =}\NormalTok{ growth, }\DataTypeTok{estimate =}\NormalTok{ growth_hat) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{pull}\NormalTok{(.estimate)}
\NormalTok{model_rmse_cv}
\CommentTok{## [1] 0.1402209}
\end{Highlighting}
\end{Shaded}

The resulting RMSE of 0.1402209 computed using cross-validation is
larger than the earlier RMSE of 0.1281398, suggesting that models that
do not take the inherent spatial autocorrelation of the data into
account generate error estimates that are overly optimistic; in our case
RMSE's that are too low.

@@

\begin{itemize}
\tightlist
\item
  show image of PLOSOne
\item
  run time considerations
\end{itemize}
